A lottery model of density-dependent selection

### 2017-10-30 manuscript.tex
latest

Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument. We therefore  briefly review that argument now. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$. The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, and led MacArthur to conclude that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}

Note that only $K_{11}$ and $K_{22}$ conform with the widespread logistic interpretation of ``$K$'' as representing saturation density; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first. To avoid this confusion we can distinguish selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}. Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.

Yet  ``$K$-selection'' is incompatible with the most widely used model of selection.

there is a deeper issue with the $r$/$K$ scheme as a description of selection in crowded versus uncrowded conditions ---

In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).

>>>>

The $r$/$K$ scheme is almost universally discussed in terms of the logistic model. This has obscured the remarkable generality of MacArthur's argument, which we thus now briefly review. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for a type to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that the other type's nullcline does not. Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth in a region of greater $n_1$ and/or $n_2$ than can be tolerated by the other type. The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, and led MacArthur to conclude that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}

Note that while $K$ is often interpreted as saturation density following the logistic model \citep{gill_1974}, MacArthur's ``$K$'' is broader: $K_{11}$ and $K_{22}$ are saturation densities while $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, where saturation density $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, and $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects. Obviously intra- and inter-type competition need not be closely connected, but in the case of exploitation competition for consumable resources they are connected via resource use ``efficiency''  \citep{gill_1974,case_1974}. Nevertheless, given the generality of the preceding argument, it seems hard to disagree that ``fitness is $K$'' in the general sense that selection at high density either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.

It is therefore remarkable that MacArthur's $K$-selection is incompatible with the standard representation of selection in evolutionary biology. This representation focuses on the change the proportion of different types (frequencies) over time,

each type is assigned a selection coefficient describing the change in its relative proportion over time (e.g. \cite[pp. 468]{barton_2007}).
